---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The PDB database

The [PDB](http://www.rcsb.org/ ) is the main repository for biomolecular structure data. 

Here we examine the contents of the PDB: 

> Q1 Download a CSV file from the PDB site (accessible from “Analyze” -> “PDB Statistics” > “by Experimental Method and Molecular Type”. Move this CSV file into your RStudio project and determine the percentage of structures solved by X-Ray and Electron Microscopy. From the website what proportion of structures are protein? Aim to have a rendered GitHub document with working code that yields your answers.

```{r}
db <- read.csv("Data Export Summary.csv", row.names=1)
head(db)
```


What percentage are X-Ray?
```{r}
db$Total[1]/(sum(db$Total)) * 100
```

What percentage are Electron Microscopy?
```{r}
db$Total[3]/(sum(db$Total)) * 100
```

What percentage are Protein?
```{r}
(sum(db$Proteins))/(sum(db$Total)) * 100
```


Using the datapasta package to copy paste any table into R without needing a .csv file
```{r}
library(datapasta)

# copy table from site, and go to add-ins and paste as data frame
# can paste any data table from any website, even if it doesn't have a .csv with it

tmp <- data.frame(stringsAsFactors=FALSE,
   Experimental.Method = c("X-Ray", "NMR", "Electron Microscopy", "Other",
                           "Multi Method", "Total"),
              Proteins = c(126880, 11062, 2277, 256, 129, 140604),
         Nucleic.Acids = c(2012, 1279, 31, 4, 5, 3331),
    ProteinComplex = c(6547, 259, 800, 6, 2, 7614),
                 Other = c(8, 8, 0, 13, 1, 30),
                 Total = c(135447, 12608, 3108, 279, 137, 151579)
)

View(tmp)

```



> Q2 Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?
There are 1157 structures as of 2019-05-07 (http://www.rcsb.org/pdb/results/results.do?tabtoshow=Current&qrid=ADAE5411)







## Now working with Bio3D and pdb 


```{r}
library(bio3d)
example(plot.bio3d)
```



Reading a pdb file
```{r}
pdb <- read.pdb("1hsg.pdb")
pdb
```






```{r}
aa321(pdb$atom$resid)
```


Atom selection is done via the function **atom.select()**

```{r}
prot.pdb <- atom.select(pdb, "protein", value=TRUE)
write.pdb(prot.pdb, file="1hsg_protein.pdb")
```


```{r}
lig.pdb <- atom.select(pdb, "ligand", value=TRUE)
write.pdb(lig.pdb, file="1hsg.ligand.pdb")
```





Section 5 of the hands on worksheet


```{r}
aa <- get.seq("1ake_A")
```

```{r}
# Blast or hmmer search
b <- blast.pdb(aa)
```

```{r}
# Plot a summary of search results
hits <- plot(b)
```


```{r}
head(hits$pdb.id)
```



```{r}
# Fetch PDBs
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE) 
# Align structures
pdbs <- pdbaln(files)
# Vector containing PDB codes
ids <- basename.pdb(pdbs$id)
# Draw schematic alignment
plot(pdbs, labels=ids)
```



```{r}
# Calculate sequence conservation
cons <- conserv(pdbs, method="entropy22") 
# SSE annotations
sse <- pdbs2sse(pdbs, ind=1, rm.gaps=FALSE)
# Plot conservation per residue
plotb3(cons, sse=sse, ylab="Sequence entropy")
```



```{r}
library(XML)
library(RCurl)
anno <- pdb.annotate(ids) 
print(unique(anno$source))
```








































