---
title: "Class 9: https://bioboot.github.io/bimm143_S19/class-material/lab-9-bimm143.html"
author: "Saeed Ghassemzadeh"
date: "4/30/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Preparing the data

```{r}
wisc.df <- read.csv("WisconsinCancer.csv")
head(wisc.df)
```


Next use as.matrix() to convert the other features (i.e. columns) of the data (in columns 3 through 32) to a matrix. Store this in a variable called wisc.data.
```{r}
wisc.data <- as.matrix(wisc.df[,3:32]) # the 33rd column is filled with NA, we don't want that
```


Assign the row names of wisc.data the values currently contained in the id column of wisc.df. While not strictly required, this will help you keep track of the different observations throughout the modeling process.
```{r}
rownames(wisc.data) <- wisc.df$id
head(wisc.data)
```


Finally, setup a separate new vector called diagnosis to be 1 if a diagnosis is malignant ("M") and 0 otherwise. Note that R coerces TRUE to 1 and FALSE to 0. 
```{r}
# Create diagnosis vector by completing the missing code
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
```


> Q1. How many patients are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))
```

> Q3. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```




# Performing PCA


Check the mean and standard deviation of the features (i.e. columns) of the wisc.data to determine if the data should be scaled. Use the colMeans() and apply() functions like you’ve done before.
```{r}
# Check column means and standard deviations
round(colMeans(wisc.data), 1) # rounding each of the means in the columns to 1 decimal 

round(apply(wisc.data,2,sd), 1) # the standard deviation along the columns (2 means columns, 1 would mean rows) rounded to 1 decimal
```


Execute PCA with the prcomp() function on the wisc.data, scaling if appropriate, and assign the output model to wisc.pr.
```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
# proportion of variance is 44% for PC1, 18% for PC2; collectively, PC1 and PC2 are 63%; PC1 and PC2 and PC3 is 72% ("cumulative")
```


> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
44%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
3 PCs

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
7 PCs


```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=diagnosis+1)
```



# Interpreting PCA results


Create a biplot of the wisc.pr using the biplot() function.
```{r}
biplot(wisc.pr)
# turns out to be useless
```


Rownames are used as the plotting character for biplots like this one which can make trends rather hard to see. In fact, this plot is very poor. So lets generate a more standard scatter plot of each observation along principal components 1 and 2 (i.e. a plot of PC1 vs PC2 available as the first two columns of wisc.pr$x) and color the points by the diagnosis (available in the diagnosis vector you created earlier).
```{r}
# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=diagnosis+1 , xlab = "PC1", ylab = "PC2")
```



Because principal component 2 explains more variance in the original data than principal component 3, you can see that the first plot has a cleaner cut separating the two subgroups.
```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[,c(1,3)], col = (diagnosis + 1), xlab = "PC1", ylab = "PC3")
```
Black points are benign. 
Red points are malignant. 


# Variance explained


Calculate the variance of each principal component by squaring the sdev component of wisc.pr (i.e. wisc.pr$sdev^2). Save the result as an object called pr.var. 
```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```



Calculate the variance explained by each principal component by dividing by the total variance explained of all principal components. Assign this to a variable called pve and create a plot of variance explained for each principal component.
```{r}
# Variance explained by each principal component: pve
pve <- pr.var/sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", ylab = "Proportion of Variance Explained", ylim = c(0, 1), type = "o")
```


```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained", names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```



Using the cumsum() function, create a plot of cumulative proportion of variance explained. 
```{r}
# Plot cumulative proportion of variance explained
plot(cumsum(pve), xlab = "Principal Component", ylab = "Cumulative Proportion of Variance Explained", ylim = c(0, 1), type = "o")
```





Use the par() function to create a side by side plot (i.e. 1 row 2 column arrangement) of these two graphs.
```{r}
par(mfcol=c(1,2))
plot(pve, xlab = "Principal Component", ylab = "Proportion of Variance Explained", ylim = c(0, 1), type = "o")
plot(cumsum(pve), xlab = "Principal Component", ylab = "Cumulative Proportion of Variance Explained", ylim = c(0, 1), type = "o")
```



```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)

fviz_eig(wisc.pr, addlabels = TRUE)
```



# Communicating PCA results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?


> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?





# Hierarchical clustering of case data



Scale the wisc.data data and assign the result to data.scaled.
```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
```

Calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset and assign the result to data.dist. 
```{r}
data.dist <- dist(data.scaled)
```

Create a hierarchical clustering model using complete linkage. Manually specify the method argument to hclust() and assign the results to wisc.hclust.
```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```



# Results of hierarchical clustering


> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```




# Selecting number of clusters


Use cutree() to cut the tree so that it has 4 clusters. Assign the output to the variable wisc.hclust.clusters.
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```


We can use the table() function to compare the cluster membership to the actual diagnoses.
```{r}
table(wisc.hclust.clusters, diagnosis)
# column 0 --> benign, 12 benign patients clustered into cancer group (misdiagnosed)
# row 3 --> 343 benign patients in benign, but also 40 cancer patients clustered in with them (misdiagnosed)
```


> Q12. Can you find a better cluster vs diagnoses match with by cutting into a different number of clusters between 2 and 10?




# Clustering on PCA results

Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with the linkage method="ward.D2". We use Ward’s criterion here because it is based on multidimensional variance like principal components analysis. Assign the results to wisc.pr.hclust.
```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2") 
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis+1)
```






# Prediction


We will use the predict() function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.
```{r}
url <- "new_samples.csv"
#url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=15)
```

> Q17. Which of these new patients should we prioritize for follow up based on your results?










































