---
title: "Class 7: R functions and packages"
author: "Saeed Ghassemzadeh"
date: "4/23/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





# Functions revisited

We will source a file from online with our functions from last day

```{r}
source("http://tinyurl.com/rescale-R")
```

Try out the last day's rescale() function

```{r}
rescale(1:10)
```

```{r}
## test fail 
## rescale(c(1:10, "string"))
```


Try the rescale2() function that catches string inputs
```{r}
rescale2(c(1:10, "string"))
```







# Find missing NA values in two vectors

Start with a simple example of the larger problem I am trying to solve

```{r}
x <- c( 1, 2, NA, 3, NA) 
y<-c(NA, 3, NA, 3, 4)
```


```{r}
is.na(x)
```

```{r}
is.na(y)
```

```{r}
which(is.na(x))
```

Try putting these together with an AND

```{r}
is.na(x) & is.na(y)
```

Take the sum() to find out how many TRUE values we have and thus how many NAs we had in both x and y

```{r}
sum(is.na(x) & is.na(y))
```

Now I can make this into our first function...

```{r}
both_na <- function(x, y) {
  sum(is.na(x) & is.na(y))
}
  
```


```{r}
both_na(x, y)
```

```{r}
both_na(x, c(NA, 3, NA, 2, NA))
```

```{r}
x <-  c(NA, NA, NA)
y1 <- c( 1, NA, NA)
y2 <- c( 1, NA, NA, NA)
```


```{r}
both_na(x, y2)
# will recycle length of shorter vector to match length of longer vector
# essentially gives x an extra NA at the end
```


```{r}
y3 <- c(1, NA, NA, NA, NA, NA, NA)
# compare with y1 (run below)
# x = match
#     (1), x,  x, (1), x,  x, (1)
# result is 4 (see below)
```

```{r}
both_na(y1, y3)
```

```{r}
3 == 3

3 < 2

3 == 2

3 != 2
```

```{r}
length(y2)
```

```{r}
length(x) != length(y2)
# see above, x has 3 things in it, y2 has 4 things in it
```

Now let's try the both_na2() function on our different length input vactors

```{r}
# both_na2(x, y2)
# there is a stop; unlike both_na(), this function STOPS the code early
```


```{r}
which(c(F,F,T,F,T))
# tells you which slot is TRUE
# can also write out FALSE or TRUE
```


```{r}
which(is.na(c(1,2,NA,4)))
# tells you which slot is an NA
```


```{r}
x <- c( 1, 2, NA, 3, NA) 
y<-c(NA, 3, NA, 3, 4)

both_na3(x,y)

# $number tells you the count of how many both NA there is 
# $which tells you which slot they both have an NA
```



# One last example

```{r}
df1
```


```{r}
df2
```

```{r}
df3
```


Make things simple

```{r}
x <- df1$IDs
y <- df2$IDs

x
y
```


```{r}
intersect(x,y)
```


```{r}
x %in% y
# which position elements in x are seen anywhere in y
```

```{r}
which(x %in% y)
# which specific position elements in x are sene in y
```


```{r}
(x %in% y)
```


```{r}
x[x %in% y]
```


```{r}
# something about "indices"...
x[x %in% y]

y[ y %in% x ]

```


# Use the R Studio shortcut 'Code -> Extract Function' to turn our snippet into a working function

```{r}
# We can now cbind() these these results...
gene_intersect <- function(x, y) {
  cbind( x[ x %in% y ], y[ y %in% x ] )
} 

# code -> extract function -> name function
# it tries to generate you a function, check it to make sure everything is right
```

```{r}
gene_intersect(df1$IDs, df2$IDs)
```



```{r}
gene_intersect2(df1, df2)
# ugly printing of column name at the end
```

```{r}
gene_intersect3(df1, df2)
# this function cleans up the column name
```



```{r}
merge(df1, df2, by="IDs")
```




---------------------------------------------------------------------------------------------
## Grade Function

Write a function called grade() to calculate the mean score, dropping the lowest single score

```{r}
x <- c(100, 100, 100, 100, 100, 100, 100, 90)
# set2 <- c(100, 90, 90, 90, 90, 90, 97, 80)
```

Find out what's the worst score. Subtract from class. Then, do average of that set. 
```{r}
min(x)

sum(x) - min(x)

(sum(x) - min(x))/(length(x)-1)
```

Create the function
```{r}
grade <- function(x) {
  (sum(x) - min(x))/(length(x)-1)
}
  
```

Test the function
```{r}
grade(set1)
```


-----------------



























































